\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-27 paper/experiments.tex
paper: pass over experiments section

\subsection{Data sets} \label{subsec:datasets}
\subsubsection{CDC NHANES}
We use survey data collected by the 
Centers for Disease Control and Prevention (CDC) in the United States. We
experiment on a set of tables collected as part of the 2013--2014 National
Health and Nutrition Examination Survey (NHANES), a series of studies
conducted by the CDC on a national sample of several thousand individuals~\cite{cdc-data}.
The data consists on survey responses, physical examinations, and laboratory
results, amongst others.

@@

\begin{table}
   \centering
    \input{tables/fcc}
    \caption{Missing values in fCC Survey Data}
   \label{table:fcc-description} 
\end{table}

@@

\subsection{Queries}
We collected a set of queries (\Cref{fig:queries}) that we think are both representative and interesting
to plan. We believe that they could reasonably be written by a user in the course of data analysis.

@@

Queries 5-8 are on the fCC data. In~\Cref{q5}, we calculate the average income for
survey participants, based on their bootcamp attendance. In~\Cref{q6} estimates the
average age of women from the United States who participated. In~\Cref{q7}, we
calculate the average amount of money survey participants with student debt destined to
learning based on their school degree. In~\Cref{q8}, we join the fCC data with a
reference table provided by the World Bank which summarizes GDP per-capita across various
countries\cite{worldbank-data}. The query calculates the average GDP per-capita for
participants based on whether or not they attended a bootcamp.

Note that for all queries we have enumerated strings and encoded them with
an appropriate integer value.

\begin{table*}
\centering
 \begin{subtable}{\linewidth}
  \input{queries-cdc}
  \caption{Queries on CDC data}
  \label{fig:queries-cdc}
 \end{subtable}
 ~
 \begin{subtable}{\linewidth}
 \input{queries-fcc}
 \caption{Queries on fCC data}
 \label{fig:queries-fcc}
 \end{subtable}
  \caption{Queries used in our experiments.}
  \label{fig:queries}
\end{table*}

@@

All experiments were run on a single Amazon Web Services EC2 {\tt c4.xlarge} instance, with
four 2.9 GHz Intel Xeon E5-2666 v3 virtual CPUs and 7.5 GiB of main memory, on Debian Linux.

@@

\begin{figure}
\todo{change this to a single graph of X-times faster than base approach}
\begin{subfigure}{\linewidth}
\includegraphics[width=\columnwidth]{figures/running_times_imputedb.png}
\caption{Query runtimes with \ProjectName{}}
\label{subfig:project-runtime-queries}
\end{subfigure}
~
\begin{subfigure}{\linewidth}
\includegraphics[width=\columnwidth]{figures/running_times_base_tables.png}
\caption{Query runtimes with imputation on base tables}
\label{subfig:base-runtime-queries}
\end{subfigure}
\caption{Comparing \ProjectName{} runtimes for each query
    (\Cref{subfig:project-runtime-queries}) with the runtime for a query when base-table
    imputations are used (\Cref{subfig:base-runtime-queries})}
\label{fig:runtimes}
\end{figure}

@@

\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for
\ProjectName{}'s query results when compared to running imputation on the base tables and
executing the query on the cleaned data. Each query ran for 200 iterations in both settings
and then results from the two approaches were paired up and compared tuple-wise. We average
tuple-wise absolute percentage deviations within each iteration of a query, and we report
this value averaged over all iterations.  We can see that optimizing for quality indeed
reduces the SMAPE of query results. In general, the SMAPE relative to the base-imputation
approach are low in all cases\todo{give actual value}, indicating that on-the-fly imputation produces similar
results to imputation at the base tables.

\begin{table}
\todo{Consider switching to graph, or changing table layout..not very intuitive right now.}
\centering
\input{tables/perf_summary.tex}
\caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$
    parameterizations relative to results when imputing on base table. Values of $0.0$,
    $100.0$, or $NaN$ indicate uninformative values.} %TODO MJS
\label{table:smape}
\end{table}

In many real-world cases, applying the imputation step at the base table is prohibitive.
To illustrate the increasing difficulty of such an approach as datasets scale, we used a simple
aggregate query over the ACS dataset in our final experiment(\Cref{query-acs}).
Applying the imputation operation to the base table is extremely expensive, in our case
completing in 75 minutes. In contrast, \ProjectName{} executes a quality-optimized version
in 7 seconds and a runtime-optimized version in 1 second. This highlights the potential
benefit of using our system for early data exploration.

\begin{figure}
\begin{lstlisting}
SELECT AVG(c0) FROM acs_dirty;
\end{lstlisting}
\caption{A simple aggregate query on ACS data. Imputing the base table is prohibitive.
    \ProjectName's quality-optimized and performance-optimized queries execute in 7 seconds
    and 1 second, respectively.
}
\label{query-acs}
\end{figure}

>>>>

\subsection{Data sets} \label{subsec:datasets}
We collected three data sets our experiments.
For all data sets, we selected a subset of the original attributes.
We also enumerated strings and encoded them with an appropriate integer value.
\todo{describe all data transformations. jose?}

\subsubsection{CDC NHANES}
For our first set of experiments, we use survey data collected by the 
Centers for Disease Control and Prevention (CDC) in the United States. We
experiment on a set of tables collected as part of the 2013--2014 National
Health and Nutrition Examination Survey (NHANES), a series of studies
conducted by the CDC on a national sample of several thousand individuals~\cite{cdc-data}.
The data consists on survey responses, physical examinations, and laboratory
results, amongst others.

@@

\begin{table}
  \centering
  \input{tables/fcc}
  \caption{Missing values in freeCodeCamp Survey Data}\label{table:fcc-description} 
\end{table}

@@

\subsection{Queries}
We collected a set of queries (\Cref{fig:queries}) that we think are interesting to plan.
We believe that they could reasonably be written by a user in the course of data analysis.

@@

Queries 5-8 are on the freeCodeCamp data.
\Cref{q5} calculates the average income for survey participants, based on their bootcamp attendance.
\Cref{q6} estimates the average age of women from the United States who participated.
\Cref{q7} calculates the average amount of money survey participants with student debt spent on learning based on their school degree.
\Cref{q8} joins the freeCodeCamp data with a reference table provided by the World Bank which summarizes GDP per-capita across various countries~\cite{worldbank-data}.
The query calculates the average GDP per-capita of countries with and without bootcamp participants.

\begin{table*}
  \todo{do we really need headers on these tables?}
  \centering
  \begin{subtable}{\linewidth}
    \input{queries-cdc}
    \caption{Queries on CDC data}\label{fig:queries-cdc}
  \end{subtable}
  \par\medskip
  \begin{subtable}{\linewidth}
    \input{queries-fcc}
    \caption{Queries on freeCodeCamp data}\label{fig:queries-fcc}
  \end{subtable}
  \par\medskip  
  \caption{Queries used in our experiments.}\label{fig:queries}
\end{table*}

@@

All experiments were run on a single Amazon Web Services EC2 {\tt c4.xlarge} instance, with
four 2.9 GHz Intel Xeon E5--2666 v3 virtual CPUs and 7.5 GiB of main memory, on Debian Linux.

@@

\begin{figure}
  \todo{change this to a single graph of X-times faster than base approach}
  \begin{subfigure}{\linewidth}
    \includegraphics[width=\columnwidth]{figures/running_times_imputedb.png}
    \caption{Query runtimes with \ProjectName{}}\label{subfig:project-runtime-queries}
  \end{subfigure}
  ~
  \begin{subfigure}{\linewidth}
    \includegraphics[width=\columnwidth]{figures/running_times_base_tables.png}
    \caption{Query runtimes with imputation on base tables}\label{subfig:base-runtime-queries}
  \end{subfigure}
  \par\medskip
  \caption{Comparing \ProjectName{} runtimes for each query
    (\Cref{subfig:project-runtime-queries}) with the runtime for a query when base-table
    imputations are used (\Cref{subfig:base-runtime-queries})}\label{fig:runtimes}
\end{figure}

@@

\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for \ProjectName{}'s query results when compared to running imputation on the base tables and executing the query on the cleaned data \todo{SMAPE citation}.
Each query ran for 200 iterations \todo{iterations? is this for the imputation or is it the number of query re-runs?} in both settings and then results from the two approaches were paired up and compared tuple-wise.
We average tuple-wise absolute percentage deviations within each iteration of a query, and we report this value averaged over all iterations.
We can see that optimizing for quality indeed reduces the SMAPE of query results.
In general, the SMAPE relative to the base-imputation approach are low in all cases\todo{give actual value}, indicating that on-the-fly imputation produces similar results to imputation at the base tables.

\begin{table}
\todo{Consider switching to graph, or changing table layout..not very intuitive right now.}
\centering
\input{tables/perf_summary.tex}
\caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$
    parameterizations relative to results when imputing on base table. Values of $0.0$,
    $100.0$, or $NaN$ indicate uninformative values\todo{use one flag value (-) or explain why value is uninformative}.} %TODO MJS
\label{table:smape}
\end{table}

In many real-world cases, applying the imputation step at the base table is prohibitively expensive.
To illustrate the increasing difficulty of such an approach as datasets scale, we ran the following query over the ACS dataset:
\begin{lstlisting}
SELECT AVG(c0) FROM acs_dirty;
\end{lstlisting}
Imputing the base table and then running the query takes 75 minutes.
In contrast, \ProjectName{} executes a quality-optimized query plan in 7 seconds and a runtime-optimized plan in 1 second.
This highlights the benefit of using our system for early data exploration.
\end{figure}
